---
title: "DATA 622 Assignment 2"
subtitle: "CUNY: Spring 2021"
author: "Philip Tanofsky"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: pdf_document
---

# Introduction

```{r warning=F, message=F}
# Import required R libraries
library(palmerpenguins)
```


```{r warning=F,message=F}
ds <- penguins

head(ds)

summary(ds)

dim(ds)

glimpse(ds)

visdat::vis_dat(ds)
```

# LDA: Linear Discrimant Analysis

http://www.sthda.com/english/articles/36-classification-methods-essentials/146-discriminant-analysis-essentials-in-r/

https://www.geeksforgeeks.org/linear-discriminant-analysis-in-r-programming/


# QDA: Quadratic Discrimant Analysis

Same link as above

# NB: Naive Bayes

https://www.r-bloggers.com/2018/01/understanding-naive-bayes-classifier-using-r/


https://www.geeksforgeeks.org/naive-bayes-classifier-in-r-programming/


# ==== Prompt =====

Homework # 2 (Generative Models) (100 points) Due on March 12, 11:59pm EST
We will be working with the Penguin dataset again as we did for Homework #1. Please use “Species” as your target variable. For this assignment, you may want to drop/ignore the variable “year”.
Using the target variable, Species, please conduct:
a. LinearDiscriminantAnalysis(30points):
a. Youwanttoevaluateallthe‘features’ordependentvariablesandsee what should be in your model. Please comment on your choices.
b. Justasuggestion:YoumightwanttoconsiderexploringfeaturePlot on the caret package. Basically, you look at each of the features/dependent variables and see how they are different based on species. Simply eye-balling this might give you an idea about which would be strong ‘classifiers’ (aka predictors).
c. Fit your LDA model using whatever predictor variables you deem appropriate. Feel free to split the data into training and test sets before fitting the model.
d. Lookatthefitstatistics/accuracyrates.
b. QuadraticDiscriminantAnalysis(30points)
a. Samestepsasabovetoconsider
c. Naïve Bayes (30 points)
a. Samestepsasabovetoconsider
d. Commentonthemodelsfits/strength/weakness/accuracyforallthesethree models that you worked with. (10 points)
