---
title: "DATA 622 Assignment 3"
subtitle: "CUNY: Spring 2021"
author: "Philip Tanofsky"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: pdf_document
---

```{r warning=F, message=F}
# Import required R libraries
library(vcd)
library(caret)
#library(MASS)
#library(ggplot2)
#library(mvtnorm)
#library(e1071)
#library(klaR)
#library(pROC)
#library(corrplot)
theme_set(theme_classic())

library(tidyverse)
library(tidymodels)
library(skimr)
library(baguette)
library(future)
library(xgboost)
library(vip)
library(rpart.plot)
```


```{r warning=F, message=F}
# Read in loan approval csv
data <- read.csv("https://raw.githubusercontent.com/ptanofsky/data622/main/assignment03/Loan_approval.csv")

data$Credit_History <- as.factor(data$Credit_History)

data$Total_Income <- data$ApplicantIncome + data$CoapplicantIncome

data$LoanAmt_Per_Month <- data$LoanAmount / data$Loan_Amount_Term

data$Income_To_LoanAmt <- data$Total_Income / data$LoanAmount

data$Income_To_LoanAmtMonth <- data$Total_Income / data$LoanAmt_Per_Month

summary(data)

dim(data)
```

Dimensions:
614 observations

13 columns

All columns factor except:

ApplicationIncome: int

CoapplicantIncome: num
LoanAmount: int
Loan_Amount_Term: int
Credit_History: int, should probably be factor

Loan_ID: Unique identifier
Gender: Female|Male
Married: No|Yes




```{r warning=F, message=F,eval=F}
# Count penguins for each loan status / gender
ggplot(data, aes(x = Gender, fill = Loan_Status)) +
  geom_bar(alpha = 0.8) +
  scale_fill_manual(values = c("darkorange", "purple", "cyan4"),
                    guide = F) +
  theme_minimal() +
  facet_wrap(~Loan_Status, ncol = 1) +
  coord_flip()

mosaic(~ Gender + Loan_Status, data = data)
```

```{r warning=F, message=F,eval=F}
# Count penguins for each loan status / married
ggplot(data, aes(x = Married, fill = Loan_Status)) +
  geom_bar(alpha = 0.8) +
  scale_fill_manual(values = c("darkorange", "purple", "cyan4"),
                    guide = F) +
  theme_minimal() +
  facet_wrap(~Loan_Status, ncol = 1) +
  coord_flip()

mosaic(~ Married + Loan_Status, data = data)
```

```{r warning=F, message=F,eval=F}
# Count penguins for each loan status / dependents
ggplot(data, aes(x = Dependents, fill = Loan_Status)) +
  geom_bar(alpha = 0.8) +
  scale_fill_manual(values = c("darkorange", "purple", "cyan4"),
                    guide = F) +
  theme_minimal() +
  facet_wrap(~Loan_Status, ncol = 1) +
  coord_flip()

mosaic(~ Dependents + Loan_Status, data = data)
```

```{r warning=F, message=F,eval=F}
# Count penguins for each loan status / Education
ggplot(data, aes(x = Education, fill = Loan_Status)) +
  geom_bar(alpha = 0.8) +
  scale_fill_manual(values = c("darkorange", "purple", "cyan4"),
                    guide = F) +
  theme_minimal() +
  facet_wrap(~Loan_Status, ncol = 1) +
  coord_flip()

mosaic(~ Education + Loan_Status, data = data)
```

```{r warning=F, message=F,eval=F}
# Count penguins for each loan status / Self_Employed
ggplot(data, aes(x = Self_Employed, fill = Loan_Status)) +
  geom_bar(alpha = 0.8) +
  scale_fill_manual(values = c("darkorange", "purple", "cyan4"),
                    guide = F) +
  theme_minimal() +
  facet_wrap(~Loan_Status, ncol = 1) +
  coord_flip()

mosaic(~ Self_Employed + Loan_Status, data = data)
```

```{r warning=F, message=F,eval=F}
# Count penguins for each loan status / Credit_History
ggplot(data, aes(x = Credit_History, fill = Loan_Status)) +
  geom_bar(alpha = 0.8) +
  scale_fill_manual(values = c("darkorange", "purple", "cyan4"),
                    guide = F) +
  theme_minimal() +
  facet_wrap(~Loan_Status, ncol = 1) +
  coord_flip()

mosaic(~ Credit_History + Loan_Status, data = data)
```

```{r warning=F, message=F,eval=F}
# Count penguins for each loan status / Property_Area
ggplot(data, aes(x = Property_Area, fill = Loan_Status)) +
  geom_bar(alpha = 0.8) +
  scale_fill_manual(values = c("darkorange", "purple", "cyan4"),
                    guide = F) +
  theme_minimal() +
  facet_wrap(~Loan_Status, ncol = 1) +
  coord_flip()

mosaic(~ Property_Area + Loan_Status, data = data)
```

```{r warning=F,message=F,eval=F}
# Overlayed density plots
featurePlot(x = data[, 7:10],
            y = data$Loan_Status,
            plot = "density",
            # Pass in options to xyplot() to
            # make it prettier
            scales = list(x = list(relation="free"),
                          y = list(relation="free")),
            adjust = 1.5,
            pch = "|",
            layout = c(2, 2),
            auto.key = list(columns = 3))

# Overlayed density plots
featurePlot(x = data[, 14:17],
            y = data$Loan_Status,
            plot = "density",
            # Pass in options to xyplot() to
            # make it prettier
            scales = list(x = list(relation="free"),
                          y = list(relation="free")),
            adjust = 1.5,
            pch = "|",
            layout = c(2, 2),
            auto.key = list(columns = 3))
```


```{r warning=F,message=F,eval=F}
# Use featurePlot
# https://topepo.github.io/caret/visualizations.html

# Scatterplot
featurePlot(x = data[, 7:10],
            y = data$Loan_Status,
            plot = "pairs",
            # Add a key at the top
            auto.key = list(columns = 3))
            
featurePlot(x = data[, 14:17],
            y = data$Loan_Status,
            plot = "pairs",
            # Add a key at the top
            auto.key = list(columns = 3))
```

```{r warning=F,message=F,eval=F}
featurePlot(x = data[, 7:10], 
            y = data$Loan_Status,
            plot = "box", 
            ## Pass in options to bwplot() 
            scales = list(y = list(relation="free"),
                          x = list(rot = 90)),  
            layout = c(2,2), 
            auto.key = list(columns = 2))

featurePlot(x = data[, 14:17], 
            y = data$Loan_Status,
            plot = "box", 
            ## Pass in options to bwplot() 
            scales = list(y = list(relation="free"),
                          x = list(rot = 90)),  
            layout = c(2,2), 
            auto.key = list(columns = 2))
```

## Decision Tree for Loan Approval data

decision_tree() function from tidymodels

3 hyperparameters
- cost_complexity
- tree_depth
- min_n

```{r warning=F,message=F}
# https://www.gmudatamining.com/lesson-13-r-tutorial.html

lap_data <- data
summary(data)


# Data splitting
set.seed(1234)

lap_data_split <- initial_split(lap_data, prop=0.75,
                              strata = Loan_Status)

lap_training <- lap_data_split %>% training()

lap_test <- lap_data_split %>% testing()

set.seed(1234)
lap_folds <- vfold_cv(lap_training, v=3)

```

```{r warning=F,message=F,evel=F}
# Data exploration
# https://bcullen.rbind.io/post/2020-06-02-tidymodels-decision-tree-learning-in-r/

# Need to fix
lap_data %>%
  select(-contains("ID")) %>%
#  modify_if(is.character, as.factor) %>%
  skim() %>%
  select()

```

```{r warning=F,message=F}
# Feature Engineering

lap_recipe <- recipe(Loan_Status ~ ., data = lap_training) %>%
  step_YeoJohnson(all_numeric(), -all_outcomes()) %>%
  step_normalize(all_numeric(), -all_outcomes()) %>%
  step_dummy(all_nominal(), -all_outcomes())

lap_recipe %>%
  prep() %>%
  bake(new_data = lap_training)

```

```{r warning=F,message=F}
# Define model
tree_model <- decision_tree(cost_complexity = tune(),
                            tree_depth = tune(),
                            min_n = tune()) %>%
  set_engine('rpart') %>%
  set_mode('classification')
```

```{r warning=F,message=F}
# Define workflow
tree_workflow <- workflow() %>%
  add_model(tree_model) %>%
  add_recipe(lap_recipe)
```

```{r warning=F,message=F}
# Create a grid of hyperparemeter values to test
tree_grid <- grid_regular(cost_complexity(),
                          tree_depth(),
                          min_n(),
                          levels = 2)
# view grid
tree_grid
```

```{r warning=F,message=F}
# Tune decision tree workflow
set.seed(1234)

tree_tuning <- tree_workflow %>%
  tune_grid(resamples = lap_folds,
            grid = tree_grid)

tree_tuning %>% show_best('roc_auc')
```

```{r warning=F,message=F}
# Select best model based on roc_auc
best_tree <- tree_tuning %>%
  select_best(metric = 'roc_auc')

# view the best tree parameters
best_tree
```

```{r warning=F,message=F}
# finalize workflow
final_tree_workflow <- tree_workflow %>%
  finalize_workflow(best_tree)
```

```{r warning=F,message=F}
# fit the model
tree_wf_fit <- final_tree_workflow %>%
  fit(data = lap_training)
```

```{r warning=F,message=F}
tree_fit <- tree_wf_fit %>%
  pull_workflow_fit()

vip(tree_fit)
```

```{r warning=F,message=F}
rpart.plot(tree_fit$fit, roundint=FALSE)
```

```{r warning=F,message=F}
# train and evaluate
tree_last_fit <- final_tree_workflow %>%
  last_fit(lap_data_split)

tree_last_fit %>% collect_metrics()
```

```{r warning=F,message=F}
tree_last_fit %>% collect_predictions() %>%
  roc_curve(truth = Loan_Status, estimate = .pred_Y) %>%
  autoplot()

tree_predictions <- tree_last_fit %>% collect_predictions()

conf_mat(tree_predictions, truth = Loan_Status, estimate = .pred_class)
```

## Gradient Boosting for Loan Approval data
```{r warning=F,message=F}
# https://bcullen.rbind.io/post/2020-06-02-tidymodels-decision-tree-learning-in-r/
# Section Boosted Trees

# Specify the model
mod_boost <- boost_tree() %>%
  set_engine("xgboost", nthreads = parallel::detectCores()) %>%
  set_mode("classification")

```

```{r warning=F,message=F}
# Create workflow
boost_workflow <- workflow() %>%
  add_recipe(lap_recipe) %>%
  add_model(mod_boost)
```

```{r warning=F,message=F}
# fit the model
boost_wf_fit <- boost_workflow %>%
  fit(data = lap_training)
```

```{r warning=F,message=F}
boost_fit <- boost_wf_fit %>%
  pull_workflow_fit()

vip(boost_fit)
```

```{r warning=F,message=F}
# train and evaluate
boost_last_fit <- boost_workflow %>%
  last_fit(lap_data_split)

boost_last_fit %>% collect_metrics()
```

```{r warning=F,message=F}
boost_last_fit %>% collect_predictions() %>%
  roc_curve(truth = Loan_Status, estimate = .pred_Y) %>%
  autoplot()

boost_predictions <- boost_last_fit %>% collect_predictions()

boost_predictions

conf_mat(boost_predictions, truth = Loan_Status, estimate = .pred_class)
```








