---
title: "DATA 622 Assignment 4"
author: "Philip Tanofsky"
date: "4/23/2021"
output: html_document
---

```{r warning=F, message=F}
# Import required R libraries
#library(vcd)
library(kernlab)
library(caret)
#library(MASS)
#library(ggplot2)
#library(mvtnorm)
#library(e1071)
#library(klaR)
#library(pROC)
#library(corrplot)
theme_set(theme_classic())

library(tidyverse)
library(tidymodels)
library(skimr)
#library(baguette)
#library(future)
#library(xgboost)
library(vip)
library(rpart.plot)
```

```{r warning=F, message=F}
# Read in loan approval csv
data <- read_csv("https://raw.githubusercontent.com/completegraph/DATA622_MACHINELEARNING/main/HW4/ADHD_data.csv")
names(data) <- make.names(names(data),unique = TRUE)

# Remove ADHD individual questions
#data <- data[, -c(5:22)]

# Remove MD individual questions
#data <- data[, -c(6:20)]

# Remove Psych.meds (too many NAs)
#data <- data[, -c(21)]

# Remove instances with missing data
#data <- na.omit(data)

#data <- data %>% 
#  mutate_if(is.numeric, as.factor)

data$Suicide <- as.factor(data$Suicide)


data <- data %>% mutate(
             fSex = case_when( Sex == 1 ~ "M", 
                               Sex == 2 ~ "F" ) ,
             fRace = case_when(
                  Race == 1 ~ "WH" ,
                  Race == 2 ~ "AF" ,
                  Race == 3 ~ "HI" ,
                  Race == 4 ~ "AS" ,
                  Race == 5 ~ "NA" ,
                  Race == 6 ~ "OT"
              ) ,
             fCO = case_when(Court.order == 0 ~ "No" ,
                             Court.order == 1 ~ "Yes") ,
             fHViol = case_when( Hx.of.Violence == 0 ~ "No",
                                 Hx.of.Violence == 1 ~ "Yes") ,
             fDCond = case_when( Disorderly.Conduct == 0 ~ "No" ,
                                 Disorderly.Conduct == 1 ~ "Yes") ,
             fSuicide  = case_when( Suicide == 0 ~ "No" ,
                                    Suicide == 1 ~ "Yes"),
             fMD.Q2  = case_when( MD.Q2 == 0 ~ "No" ,
                                  MD.Q2 == 1 ~ "Yes")
             )

data$fSex <- as.factor(data$fSex)
data$fSuicide <- as.factor(data$fSuicide)
data$fHViol <- as.factor(data$fHViol)
data$fAbuse <- factor(as.character(data$Abuse), levels=c("0","1","2","3","4","5","6","7"), ordered=TRUE)
data$ADHD.Q16 <- factor(as.character(data$ADHD.Q16), levels=c("0","1","2","3","4"), ordered=TRUE)
#data$MD.Q3 <- factor(as.character(data$MD.Q3), levels=c("0","1","2","3"), ordered=TRUE)
data$fEducation <- factor(as.character(data$Education), levels=c("6","7","8","9","10","11","12","13","14","15","16","17","18","19"), ordered=TRUE)

data$MD.Q1.Count <- 0
data$MD.Q1.Count <- ifelse(data$MD.Q1a == 1, data$MD.Q1.Count+1, data$MD.Q1.Count)
data$MD.Q1.Count <- ifelse(data$MD.Q1b == 1, data$MD.Q1.Count+1, data$MD.Q1.Count)
data$MD.Q1.Count <- ifelse(data$MD.Q1c == 1, data$MD.Q1.Count+1, data$MD.Q1.Count)
data$MD.Q1.Count <- ifelse(data$MD.Q1d == 1, data$MD.Q1.Count+1, data$MD.Q1.Count)
data$MD.Q1.Count <- ifelse(data$MD.Q1e == 1, data$MD.Q1.Count+1, data$MD.Q1.Count)
data$MD.Q1.Count <- ifelse(data$MD.Q1f == 1, data$MD.Q1.Count+1, data$MD.Q1.Count)
data$MD.Q1.Count <- ifelse(data$MD.Q1g == 1, data$MD.Q1.Count+1, data$MD.Q1.Count)
data$MD.Q1.Count <- ifelse(data$MD.Q1h == 1, data$MD.Q1.Count+1, data$MD.Q1.Count)
data$MD.Q1.Count <- ifelse(data$MD.Q1i == 1, data$MD.Q1.Count+1, data$MD.Q1.Count)
data$MD.Q1.Count <- ifelse(data$MD.Q1j == 1, data$MD.Q1.Count+1, data$MD.Q1.Count)
data$MD.Q1.Count <- ifelse(data$MD.Q1k == 1, data$MD.Q1.Count+1, data$MD.Q1.Count)
data$MD.Q1.Count <- ifelse(data$MD.Q1L == 1, data$MD.Q1.Count+1, data$MD.Q1.Count)
data$MD.Q1.Count <- ifelse(data$MD.Q1m == 1, data$MD.Q1.Count+1, data$MD.Q1.Count)
#data$MD <- ifelse(data$MD.Q1.Count>= 7 & data$MD.Q2 == 1 & data$MD.Q3 >= 2, "Y", "N")
#data$MD <- as.factor(data$MD)

# Too many missing values, just remove the column
data$Psych.meds <- NULL

# Remove instances with missing data
data <- na.omit(data)

# Display skim summary
skim(data)
```
```{r warning=F, message=F, eval=F}
ggplot(data, aes(x=data$Psych.meds, color=Suicide)) +
  geom_histogram(fill="black", alpha=0.5, position="dodge")
```

```{r warning=F, message=F}
data %>%
  group_by(Suicide) %>%
  mutate(count = n())
# of the 142, Suicide Yes=45, Suicide No=97
```

http://www.rebeccabarter.com/blog/2020-03-25_machine_learning/
```{r warning=F, message=F}
# Split into training and test data
set.seed(1234)
# split the data into training (75%) and testing (25%)
data_split <- initial_split(data, prop = .75)
data_split

data_train <- training(data_split)
data_test <- testing(data_split)

data_train
```

```{r warning=F, message=F}
# Create CV object from training data
data_cv <- vfold_cv(data_train, v=5, repeats=3)
#data_cv <- vfold_cv(data_train)
```

```{r warning=F, message=F, eval=F}
# Recipe attempt v0 ... ignore for now
# Define the recipe
data_recipe <- recipe(Suicide ~ THC + Cocaine + Opioids, data = data) %>%
  step_normalize(all_numeric()) %>%
  step_knnimpute(all_predictors())
```

Age + Sex + Race + ADHD.Total + MD.TOTAL + Alcohol + THC + Cocaine + Stimulants + Sedative.hypnotics + Opioids + Court.order + Education + Hx.of.Violence + Disorderly.Conduct + Abuse + Non.subst.Dx + Subst.Dx

```{r warning=F, message=F}
# https://www.tidymodels.org/learn/work/tune-svm/
#data_recipe <- recipe(fSuicide ~ fSex + ADHD.Total + MD.TOTAL + fAbuse, data = data) %>%
# 69%
#data_recipe <- recipe(fSuicide ~ fSex + ADHD.Total + MD.Q1.Count + MD.Q2 + MD.Q3 + fAbuse, data = data) %>%
# 71%
#data_recipe <- recipe(fSuicide ~ fSex + MD.Q1.Count + MD.Q2 + MD.Q3 + fAbuse, data = data) %>%
# 0.7142857
#data_recipe <- recipe(fSuicide ~ fSex + MD.Q1.Count + fMD.Q2 + MD.Q3 + fAbuse + ADHD.Q16, data = data) %>%
# 0.6571429
#data_recipe <- recipe(fSuicide ~ fSex + MD.Q1.Count + fMD.Q2 + MD.Q3 + fAbuse, data = data) %>%
#0.6285714
#data_recipe <- recipe(fSuicide ~ fSex + MD.Q1.Count + MD.Q2 + MD.Q3 + fAbuse, data = data) %>%
#0.6285714
data_recipe <- recipe(fSuicide ~ fSex + MD.Q1.Count + MD.Q2 + MD.Q3 + fAbuse, data = data) %>%
  step_normalize(all_numeric(), -all_outcomes()) %>%
  step_dummy(all_nominal(), -all_outcomes())  %>%
  # Remove any zero variance predictors
  step_zv(all_predictors()) %>%
  # Remove any linear combinations
  step_lincomb(all_numeric())
```

From Alex: predictors for suicide
ADHD, Mood, Abuse and seem to impact women more than men.

All predictors:
Age + Sex + Race + ADHD.Total + MD.TOTAL + Alcohol + THC + Cocaine + Stimulants + Sedative.hypnotics + Opioids + Court.order + Education + Hx.of.Violence + Disorderly.Conduct + Abuse + Non.subst.Dx + Subst.Dx

```{r warning=F, message=F}
set.seed(1234)
data_rs <- bootstraps(data, times = 5)
```

```{r warning=F, message=F}
# Consider log-likelihood statistic
# using ROC curve for now
roc_vals <- metric_set(roc_auc, accuracy, kap)
```

```{r warning=F, message=F}
# Verbose turned off, save out-of-sample predictions is turned on
ctrl <- control_grid(verbose=F, save_pred=T)
```


```{r warning=F, message=F}
# SVM Model with Polynomial
svm_model <- svm_poly(cost = tune(),
                      degree = tune()) %>% 
  set_engine("kernlab") %>% 
  set_mode("classification") %>% 
  translate()
```

```{r warning=F, message=F, eval=F}
# SVM Model with Radial Basis Function
svm_model <- svm_rbf(cost = tune(),
                     rbf_sigma = tune()) %>%
  set_mode("classification") %>%
  set_engine("kernlab") %>% 
  translate()
```

```{r warning=F, message=F}
# Generate formula
formula_res <- svm_model %>%
  tune_grid(data_recipe,
    resamples = data_cv,
    metrics = roc_vals,
    control = ctrl
  )

formula_res
```

Suicide ~ Age + Sex + Race + ADHD.Total + MD.TOTAL + Alcohol + THC + Cocaine + Stimulants + Sedative.hypnotics + Opioids + Court.order + Education + Hx.of.Violence + Disorderly.Conduct + Abuse + Non.subst.Dx + Subst.Dx

```{r warning=F, message=F}
# The .metrics column contains tibbles of the performance metrics for each tuning parameter combination
formula_res %>%
  select(.metrics) %>%
  slice(1) %>%
  pull(1)
```

```{r warning=F, message=F}
# Get final resampling estimates
estimates <- collect_metrics(formula_res)
estimates
```

```{r warning=F, message=F}
# Display top combinations
show_best(formula_res, metric = "roc_auc")
```

```{r warning=F, message=F}
# Execute with a recipe
set.seed(1234)
recipe_res <-svm_model %>%
  tune_grid(data_recipe,
    resamples = data_cv,
    metrics = roc_vals,
    control = ctrl
  )
recipe_res
```

```{r warning=F, message=F}
# Best setting
show_best(recipe_res, metric = "roc_auc")
```

```{r warning=F, message=F}
# As save_pred = TRUE, we have kept the out-of-sample predictions for each sample
collect_predictions(recipe_res)
```

```{r warning=F, message=F}
# Set the workflow
svm_wf <- workflow() %>%
  add_recipe(data_recipe) %>%
  add_model(svm_model)
```

```{r warning=F, message=F, eval=F}
# Tune parameters
svm_grid <- expand.grid(mtry = c(3,4,5))

svm_tune_results <- svm_wf %>%
  tune_grid(resamples = data_cv,
            grid = svm_grid,
            metrics = metric_set(accuracy, roc_auc, kap)
            )

# Print results
svm_tune_results %>%
  collect_metrics()
```

```{r warning=F, message=F}
# Select best model based on roc_auc
best_svm <- recipe_res %>%
  select_best(metric = 'roc_auc')

# view the best svm parameters
best_svm
```

```{r warning=F,message=F}
# finalize workflow
final_svm_wf <- svm_wf %>%
  finalize_workflow(best_svm)
```

```{r warning=F,message=F}
# fit the model
svm_wf_fit <- final_svm_wf %>%
  fit(data = data_train)

#https://stackoverflow.com/questions/62772397/integration-of-variable-importance-plots-within-the-tidy-modelling-framework
svm_wf_fit
```

```{r warning=F,message=F}
library(vip)
svm_fit <- svm_wf_fit %>%
  pull_workflow_fit()

#http://www.rebeccabarter.com/blog/2020-03-25_machine_learning/#evaluate-the-model-on-the-test-set
svm_fit$fit

# Received error model-specific variable importance scores are currently not available for this type of model
#vip(svm_fit)

# https://stackoverflow.com/questions/62772397/integration-of-variable-importance-plots-within-the-tidy-modelling-framework
svm_wf_fit %>%
  pull_workflow_fit() %>%
  vip(method = "permute", 
      target = "fSuicide", metric = "rsquared",
      pred_wrapper = kernlab::predict, train = data_train)
```

```{r warning=F,message=F}
# Received error: Not an rpart object
#rpart.plot(svm_fit$fit, roundint=FALSE)
```

```{r warning=F,message=F}
# train and evaluate
svm_last_fit <- final_svm_wf %>%
  last_fit(data_split)

svm_last_fit %>% collect_metrics()

svm_last_fit
```


```{r warning=F,message=F}
svm_last_fit %>% collect_predictions() %>%
  roc_curve(truth = fSuicide, estimate = .pred_No) %>%
  autoplot()

svm_predictions <- svm_last_fit %>% collect_predictions()

conf_mat(svm_predictions, truth = fSuicide, estimate = .pred_class)

svm_predictions

mean(svm_predictions$.pred_class == svm_predictions$fSuicide)
```



```{r warning=F, message=F}
svm_predictions %>%
  ggplot() +
  geom_density(aes(x = .pred_Yes, fill = fSuicide),
               alpha = 0.5)
```

```{r warning=F, message=F}
# https://www.tidymodels.org/learn/work/tune-svm/
augment(recipe_res) %>%
  ggplot(aes(MD.Q1.Count, .pred_Yes, color = fSuicide)) +
  geom_point(show.legend = FALSE) +
  facet_wrap(~fSuicide)
```


Results

With svm_poly, no tuning
##           Truth
## Prediction  0  1
##          0 20 11
##          1  1  3

Results

with svm_rbf, with tuning
##           Truth
## Prediction  0  1
##          0 21 14
##          1  0  0

With svm_poly, with tuning
## Prediction  0  1
##          0 21 13
##          1  0  1

With svm_poly, with tuning
##           Truth
## Prediction  0  1
##          0 18 11
##          1  2  4

svm_rbf has never done anything but always predict No for suicide, no discernment


0.6904762
poly with BoxCox


https://stackoverflow.com/questions/62772397/integration-of-variable-importance-plots-within-the-tidy-modelling-framework
https://github.com/tidymodels/parsnip/issues/311
https://www.tidymodels.org/learn/work/tune-svm/
https://stackoverflow.com/questions/8287344/cannot-plot-graph-for-an-svm-model-in-r
https://stackoverflow.com/questions/23613952/support-vector-machine-visualization-in-r
https://stackoverflow.com/questions/1142294/how-do-i-plot-a-classification-graph-of-a-svm-in-r