---
title: "Data 622 Homework 4: Mental Health Data Modeling"
author: "Group 6: Alexander Ng, Scott Reed, Philip Tanofsky, Randall Thompson"
date: "Submitted by 05/07/2021"
output:
  html_document:
    df_print: paged
    highlight: pygments
    number_sections: yes
    theme: readable
    toc: yes
    toc_depth: 2
    toc_float: no
    fontsize: 12
  pdf_document:
    toc: yes
    toc_depth: '3'
    number_sections: true
    df_print: kable
    highlight: tango
editor_options:
  chunk_output_type: inline
fontsize: 11pt
urlcolor: blue
---

# Prompt

For this assignment, we will be working with a very interesting mental health dataset from a real-life research project. All identifying information, of course, has been removed. The attached spreadsheet has the data (the tab name "Data"). The data dictionary is given in the second tab. You can get as creative as you want. The assignment is designed to really get you to think about how you could use different methods.

1. Please use a clustering method to find clusters of patients here. Whether you choose to use k-means clustering or hierarchical clustering is up to you as long as you reason through your work. Can you come up with creative names for the profiles you found? (60)

2. Let's explore using Principal Component Analysis on this dataset. You will note that there are different types of questions in the dataset: column: E-W: ADHD self-report; column X – AM: mood disorders questionnaire, column AN-AS: Individual Substance Misuse; etc. Please reason through your work as you decide on which sets of variables you want to use to conduct Principal Component Analysis. (60)

3. Assume you are modeling whether a patient attempted suicide (columnAX). Please use support vector machine to model this. You might want to consider reducing the number of variables or somehow use extracted information from the variables. This can be a really fun modeling task! (80)

```{r setup, include=FALSE}
# Your libraries go here

library(tidyverse)
library(ggplot2)
library(knitr)
library(kableExtra)

library(caret)    # Model Framework
library(skimr)    # Used for EDA
library(klaR)     # Implemented KNN and Naive Bayes models, etc
library(class)    # used for KNN classifier

# PLEASE ADD YOUR R LIBRARIES BELOW
# ------------------------------
library(stats)
library(cluster)    # clustering algorithms
library(factoextra) # clustering algorithms & visualization
library(tidymodels)
library(e1071)
library(vcd)

# ---------------------------------
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)

```

## Read data in

```{r}
adhd_data <- read_csv("https://raw.githubusercontent.com/completegraph/DATA622_MACHINELEARNING/main/HW4/ADHD_data.csv")
names(adhd_data) <- make.names(names(adhd_data),unique = TRUE)

adhd_data <- adhd_data %>% 
  mutate_if(is.numeric, as.factor) 

# Remove ADHD individual questions
adhd_data <- adhd_data[, -c(5:22)]

# Remove MD individual questions
adhd_data <- adhd_data[, -c(6:20)]

# Remove Psych.meds (too many NAs)
adhd_data <- adhd_data[, -c(21)]

# Remove Initial
#adhd_data <- adhd_data[, -c(1)]


#adhd_data$Age <- as.numeric(adhd_data$Age)
#adhd_data$ADHD.Total <- as.numeric(adhd_data$ADHD.Total)
#adhd_data$MD.TOTAL <- as.numeric(adhd_data$MD.TOTAL)


#adhd_data$Age <- factor(adhd_data$Age, levels=1:41, ordered=T)
#adhd_data$ADHD.Total <- factor(adhd_data$ADHD.Total, levels=1:62, ordered=T)
#adhd_data$MD.TOTAL <- factor(adhd_data$MD.TOTAL, levels=1:18, ordered=T)
#adhd_data$Alcohol <- factor(adhd_data$Alcohol, levels=0:3, ordered=T)
#adhd_data$THC <- factor(adhd_data$THC, levels=0:3, ordered=T)
#adhd_data$Cocaine <- factor(adhd_data$Cocaine, levels=0:3, ordered=T)
#adhd_data$Stimulants <- factor(adhd_data$Stimulants, levels=0:3, ordered=T)
#adhd_data$Sedative.hypnotics <- factor(adhd_data$Sedative.hypnotics, levels=0:3, ordered=T)
#adhd_data$Opioids <- factor(adhd_data$Opioids, levels=0:3, ordered=T)
#adhd_data$Education <- factor(adhd_data$Education, levels=6:19, ordered=T)
#adhd_data$Abuse <- factor(adhd_data$Abuse, levels=0:7, ordered=T)
#adhd_data$Non.subst.Dx <- factor(adhd_data$Non.subst.Dx, levels=0:2, ordered=T)
#adhd_data$Subst.Dx <- factor(adhd_data$Subst.Dx, levels=0:3, ordered=T)

# Remove instances with missing data
adhd_data <- na.omit(adhd_data)

adhd_data <- dplyr::mutate(adhd_data, ID = row_number())

adhd_data <- adhd_data %>% relocate(ID)

# Scale/standardize the data
# Yes, this is being applied to categorical data
#adhd_data[,2:20] <- scale(adhd_data[,2:20])

dim(adhd_data)
#head(adhd_data)
#summary(adhd_data)


p1 <- ggplot(adhd_data) + geom_bar(aes(x = ADHD.Total))

p1 + facet_wrap(~ Abuse)
```

```{r warning=F, message=F}
# Count for each Abuse / suicide attempt
ggplot(adhd_data, aes(x = Abuse, fill = Suicide)) +
  geom_bar(alpha = 0.8) +
  scale_fill_manual(values = c("darkorange", "purple", "cyan4"),
                    guide = F) +
  theme_minimal() +
  facet_wrap(~Suicide, ncol = 1) +
  coord_flip()

mosaic(~ Abuse + Suicide, data = adhd_data)
```

```{r warning=F, message=F}
# Count for each Education / suicide attempt
ggplot(adhd_data, aes(x = Education, fill = Suicide)) +
  geom_bar(alpha = 0.8) +
  scale_fill_manual(values = c("darkorange", "purple", "cyan4"),
                    guide = F) +
  theme_minimal() +
  facet_wrap(~Suicide, ncol = 1) +
  coord_flip()

mosaic(~ Education + Suicide, data = adhd_data)
```

```{r warning=F, message=F}
# Count for each Subst.Dx / suicide attempt
ggplot(adhd_data, aes(x = Subst.Dx, fill = Suicide)) +
  geom_bar(alpha = 0.8) +
  scale_fill_manual(values = c("darkorange", "purple", "cyan4"),
                    guide = F) +
  theme_minimal() +
  facet_wrap(~Suicide, ncol = 1) +
  coord_flip()

mosaic(~ Subst.Dx + Suicide, data = adhd_data)
```

```{r warning=F, message=F}
# Count for each Non.subst.Dx / suicide attempt
ggplot(adhd_data, aes(x = Non.subst.Dx, fill = Suicide)) +
  geom_bar(alpha = 0.8) +
  scale_fill_manual(values = c("darkorange", "purple", "cyan4"),
                    guide = F) +
  theme_minimal() +
  facet_wrap(~Suicide, ncol = 1) +
  coord_flip()

mosaic(~ Non.subst.Dx + Suicide, data = adhd_data)
```

```{r warning=F, message=F}
# Count for each Abuse / Subst.Dx
ggplot(adhd_data, aes(x = Abuse, fill = Subst.Dx)) +
  geom_bar(alpha = 0.8) +
  scale_fill_manual(values = c("darkorange", "purple", "cyan4", "black"),
                    guide = F) +
  theme_minimal() +
  facet_wrap(~Subst.Dx, ncol = 1) +
  coord_flip()

mosaic(~ Abuse + Subst.Dx, data = adhd_data)
```

# Section 1: Clustering method
```{r eval=F}
# From: https://www.r-bloggers.com/2016/01/hierarchical-clustering-in-r-2/

clusters <- hclust(dist(adhd_data[, 2:20]))
plot(clusters)
```

```{r eval=F}
clusterCut <- cutree(clusters, 2)
```

```{r eval=F}
table(clusterCut, adhd_data$Suicide)
```

```{r}
# Not using below
# https://datascienceplus.com/k-means-clustering-in-r/
#set.seed(20)
#adhdCluster <- kmeans(adhd_data[, 2:20], 2, nstart = 20)
#adhdCluster
#table(adhdCluster$cluster, adhd_data$Suicide)

```

```{r eval=F}
# All variables must be numeric

# https://uc-r.github.io/kmeans_clustering
distance <- get_dist(adhd_data)
fviz_dist(distance, gradient = list(low = "#00AFBB", mid = "white", high = "#FC4E07"))

k2 <- kmeans(adhd_data[,2:21], centers=2, nstart=25)
#str(k2)
k3 <- kmeans(adhd_data[,2:21], centers=3, nstart=25)
k4 <- kmeans(adhd_data[,2:21], centers=4, nstart=25)
k5 <- kmeans(adhd_data[,2:21], centers=5, nstart=25)
k10 <- kmeans(adhd_data[,2:21], centers=10, nstart=25)

# plots to compare
p1 <- fviz_cluster(k2, geom = "point", data = adhd_data[,2:21]) + ggtitle("k = 2")
p2 <- fviz_cluster(k3, geom = "point", data = adhd_data[,2:21]) + ggtitle("k = 3")
p3 <- fviz_cluster(k4, geom = "point", data = adhd_data[,2:21]) + ggtitle("k = 4")
p4 <- fviz_cluster(k5, geom = "point", data = adhd_data[,2:21]) + ggtitle("k = 5")
p10 <- fviz_cluster(k10, geom = "point", data = adhd_data[,2:21]) + ggtitle("k = 10")


library(gridExtra)
grid.arrange(p1, p2, p3, p4, p10, nrow = 3)

set.seed(123)

fviz_nbclust(adhd_data[,2:21], kmeans, method = "wss")

fviz_nbclust(adhd_data[,2:21], kmeans, method = "silhouette")

# compute gap statistic
gap_stat <- clusGap(adhd_data[,2:21], FUN = kmeans, nstart = 25,
                    K.max = 10, B = 50)
# Print the result
print(gap_stat, method = "firstmax")

fviz_gap_stat(gap_stat)
```

```{r}
# https://www.r-bloggers.com/2016/06/clustering-mixed-data-types-in-r/

# Remove ID and Initial before clustering
gower_dist <- daisy(adhd_data[, 6:21],
                    metric = "gower",
                    type = list(logratio = 3))
# Check attributes to ensure the correct methods are being used
# (I = interval, N = nominal)
# Note that despite logratio being called, 
# the type remains coded as "I"
summary(gower_dist)

gower_mat <- as.matrix(gower_dist)

adhd_data[
  which(gower_mat == min(gower_mat[gower_mat != min(gower_mat)]),
        arr.ind = TRUE)[1, ], ]

adhd_data[
  which(gower_mat == max(gower_mat[gower_mat != max(gower_mat)]),
        arr.ind = TRUE)[1, ], ]


# Calculate silhouette width for many k using PAM
sil_width <- c(NA)

for(i in 2:10){
  
  pam_fit <- pam(gower_dist,
                 diss = TRUE,
                 k = i)
  
  sil_width[i] <- pam_fit$silinfo$avg.width
  
}

# Plot sihouette width (higher is better)

plot(1:10, sil_width,
     xlab = "Number of clusters",
     ylab = "Silhouette Width")
lines(1:10, sil_width)

pam_fit <- pam(gower_dist, diss = TRUE, k = 7)

pam_results <- adhd_data %>%
  dplyr::select(6:21) %>%
  mutate(cluster = pam_fit$clustering) %>%
  group_by(cluster) %>%
  do(the_summary = summary(.))

pam_results$the_summary

adhd_data[pam_fit$medoids, ]


library(Rtsne)
tsne_obj <- Rtsne(gower_dist, is_distance = TRUE, perplexity=1)

tsne_data <- tsne_obj$Y %>%
  data.frame() %>%
  setNames(c("X", "Y")) %>%
  mutate(cluster = factor(pam_fit$clustering),
         name = adhd_data$Initial)

ggplot(aes(x = X, y = Y), data = tsne_data) +
  geom_point(aes(color = cluster))
```


# Section 3: SVM


```{r}
# http://uc-r.github.io/svm


# https://www.geeksforgeeks.org/classifying-data-using-support-vector-machinessvms-in-r/

adhd_data_split <- initial_split(adhd_data, prop=0.7, strata=Suicide)

adhd_training <- adhd_data_split %>% training()

adhd_test <- adhd_data_split %>% testing()

classifier <- svm(formula = Suicide ~ .,
                 data = adhd_training[,6:21],
                 type = 'C-classification',
                 kernel = 'linear')

classifier


# Predicting the Test set results
y_pred <- predict(classifier, newdata = adhd_test[,6:21])

cm <- table(adhd_test$Suicide, y_pred)

cm


```











